On Optimality of Finite Element Integration

### 2016-01-16 TOMS_Paper/assembly.tex
again fixes here and there

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, consider the tensor operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$. To schedule sub-expressions of this kind, we have two options:
\begin{enumerate}[label=(\roman*)]
\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$. In general, however, the optimal choice depends on multiple factors: the loop size, the increase in operation count due to expansion, and the gain in code motion. We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii). Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}. This observation is fundamental for understanding the sharing elimination algorithm.

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. The algorithm is meant to be applied to each statement appearing in the independent basic block of a finite element integration loop nest. Section~\ref{sec:proof} will elucidate the claims of quasi-optimality and optimality.

The algorithm makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured tensor operations.
\end{itemize}

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree with recursive collection of identical sub-expressions. \\ 
\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.

@@

\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^0 + ... + ...T^n)$}. 
\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii).
\end{enumerate}
Go back to step (2).

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
\textit{Note: this pass speculatively assumes that structured tensor operations are in place. If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be exposed.}
\end{enumerate}
\end{Algo}

@@

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, the bilinear form arising from a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so strategy (i) is applied to both partitions (step 3). We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{S} = \mathbb{T}$.

@@

Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. We make use of a result -- the foundation of tensor contraction mode -- to simplify our task. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces. A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors. In our model, a reference tensor is simply represented by a particular sub-expression independent of $L_e$. This leads to the following algorithm.

\begin{Algo}[Pre-evaluation]
\label{algo:pre-evaluation}
\normalfont
Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$. We dissect the normal form input expression into distinct sub-expressions, each of them representing a specific monomial. Each sub-expression is then factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms. This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py}}, as a consequence of the results in~\citeN{Kirby:TC}. These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is computed by means of symbolic execution. Finally, $L_i$ is removed from $\Lambda$. 
\end{Algo}

The pre-evaluation of a monomial introduces some critical issues:
\begin{enumerate}
\item Depending on the complexity of a monomial, a certain number $t$ of temporary variables is required if pre-evaluation is performed. Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms). For certain values of ${<}t, n, S{>}$, pre-evaluation might then shift the loop nest from a CPU-bound to a memory-bound regime, which might be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count). This could outweigh the gain due to pre-evaluation.
\item A strategy for coordinating sharing elimination and pre-evaluation is needed. We observe that sharing elimination inhibits pre-evaluation, whereas pre-evaluation could expose further sharing elimination opportunities.
\end{enumerate}

@@

This observation that, for certain forms and discretizations, hoisting may bring the working set to exceed the size of some level of local memory (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU) led us to formulating the following \textit{memory constraints}.

@@

Note that Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects. We will carefully analyze both theoretical and practical implications of these memory constraints.

\subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
\label{sec:completeness}
We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization. Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression. Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest.

Sharing elimination and pre-evaluation tackle the problem of minimizing spatial and temporal redundancy given a set of memory constraints in finite element integration loop nests, using \textit{a very specific set of operators}. This needs be emphasized since, theoretically, one could find an even lower operation count by exploiting domain-specific properties, such as redundancies in basis functions.

@@

The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ is to apply the actual transformations and count the number of operations. If, on one hand, this is tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}. We then seek an analytic way of determining $\theta_{pre}$.

The first step consists of estimating the \textit{increase factor}, $\iota$. This number captures the increase in arithmetic complexity due to the transformations enabling pre-evaluation. To contextualize, consider the example in Figure~\ref{code:increase_factor}. One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix. The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $L_i$-dependent quantities. By product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.

@@

In general, however, determining $\iota$ is not so straightforward since redundant tabulations may result from common sub-expressions. Take the previous example. One can add one coefficient in the same function space as $f$, repeat the expansion, and observe that multiple sub-expressions (e.g., $b_{10}*b_{01}*...$ and $b_{01}*b_{10}*...$) will reduce to identical tables. To evaluate $\iota$, we then use combinatorics. We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product; (ii) $n$ is the number of unique basis functions involved in the expansion. In the original example, we have $n=3$ (for $b_{i0}$, $b_{i1}$, and $b_{i2}$) and $k=1$, which confirms what we intuitively found before, namely $\iota=3$. In the modified example there are two coefficients, so $k=2$, which means $\iota=6$.

If $\iota \geq I$ (the extent of the reduction loop), we already know that pre-evaluation will not be profitable. Intuitively, this means that we are introducing more operations than we are saving from pre-evaluating $L_i$.

If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. Consider again the mass matrix operator in Figure~\ref{code:increase_factor}. The corresponding monomial is characterized by the dot product of test and trial functions, so trivially $\rho = 1$.  If we instead take the example in Figure~\ref{code:poisson}, we have that $\rho = 3$ after a suitable factorization of basis functions. In general, therefore, $\rho$ depends on both form and discretization employed. To determine this parameter, we re-factorize the expression as per Algorithm~\ref{algo:pre-evaluation} and simply count the terms amenable to pre-evaluation.

\subsection{Formalization}
\label{sec:proof}
The following proposition states that our approach guarantees quasi-optimality. The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.

\begin{Prop}
\label{prop:optimal-approach}
Consider a multilinear form comprising a set of monomials $M$, and let $\Lambda$ be the corresponding finite element integration loop nest. Let $\Gamma$ be the transformation algorithm. Let $X$ be the set of pre-evaluated monomials, and let $Y = M \setminus X$. Assume that the pre-evaluation of different monomials does not result in identical tables. Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and satisfies Constraints~\ref{const:Le} and~\ref{const:TH}.
\end{Prop}
\begin{proof}
We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ and $Y$ can actually be constructed correctly.

Let $c_\Lambda$ denote the operation count for $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}). We need to show that there is no other synthesis $\Lambda''_{I}$ satisfying Constraints~\ref{const:Le} and~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$. This holds if and only if
\begin{enumerate}
\item \textit{The coordination of pre-evaluation with sharing elimination is optimal}. This boils down to prove that
\begin{enumerate}
\item \textit{pre-evaluating any $m \in Y$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
\item \textit{not pre-evaluating any $m \in X$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
\end{enumerate}
\item \textit{Sharing elimination produces quasi-optimal (or optimal) loop nests.}\\
\end{enumerate}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum. This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum that also preserves the memory constraints. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished. Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of loop linearity and normal form of expressions, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols. These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration. Since (i) is side effects free (i.e., neither expansion nor factorization are needed) and is applied if and only if $|P| < |S_P|$, it is clear that step (3) cannot prune any optimal solution from the search space. %; in fact, it simply shifts towards the objective.

>>>>

\subsubsection{Structured tensor operations}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy is to be searched in a space of size 2. To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. One of the resulting sub-expressions will be $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i)$, in which $x^0_i$ and $x^1_i$ represent components of $x$ depending on loop $L_i$. To schedule sub-expressions of this kind, we have two options:
\begin{enumerate}[label=(\roman*)]
\item Using generalized code motion -- not useful in the example as $x$ only depends on $L_i$.
\item Searching for spatial sharing, expanding, and then factorizing the expression to promote temporal sharing -- the expression in the example would be recast as $x^0_i (a + c) + x^1_i (b + d)$, with $(a + c)$ and $(b + d)$ being two hoistable sub-expressions.
\end{enumerate}
The reader can verify that (ii) improves the operation count over (i) independently of the size of $L_i$. In general, however, the optimal choice depends on multiple factors: the loop(s) size, the increase in operation count due to expansion, and the gain in code motion. We regard as \textit{structured} any tensor operation producing expressions for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of alternatives (i) and (ii). Of notable importance, from the perspective of achieving quasi-optimality, is the fact that all tensor operations involving test and trial functions are structured\footnote{It is also remarkable that structured tensor operations often characterize the \textit{entire} variational form. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}. This observation is fundamental for understanding the sharing elimination algorithm.

\subsubsection{Algorithm}
\label{sec:se-algo}
Algorithm~\ref{algo:sharing-elimination} describes sharing elimination assuming as input a tree representation of the loop nest. It makes use of the following notation and terminology:
\begin{itemize}
\item $|.|$: a generic ``sizeof'' operator (e.g., cardinality of a set, extent of a loop).
\item \textit{multilinear operand}: any $\alpha_{eij}$ or $\beta_{eik}$ in the input expression.
\item \textit{multilinear symbol}: a symbol appearing within a multilinear operand depending on $L_j$ or $L_k$ (e.g., test functions, first order derivatives of test functions, etc.).
\item \textit{strategy (i)} and \textit{strategy (ii)}: the two approaches described in Section~\ref{sec:se-rln} for handling structured tensor operations.
\end{itemize}

\begin{Algo}[Sharing elimination]
\label{algo:sharing-elimination}
\normalfont
The algorithm has three main phases: initialization (step 1); scheduling of multilinear operands for reaching quasi-optimality (steps 2-5); scheduling of all other sub-expressions (step 6).
\begin{enumerate}
\item Perform a depth-first visit of the expression tree to factorize identical terms within additions. \\ 
\textit{Note: for example, an expression $(a + a + b + c + a)$ would be transformed into $(3a + b + c)$}.

@@

\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply strategy (ii); $\mathbb{S} = \mathbb{S} \setminus \lbrace T^1, ..., T^n \rbrace$.\\ \textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}. 
\item Only one terminal $T$ adjacent to $S$: apply strategy (ii) including one non-terminal symbol $S^{nt}$. $\mathbb{S} = \mathbb{S} \setminus \lbrace T, S^{nt} \rbrace$.
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply strategy (ii).
\end{enumerate}
Go back to step (2).

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
\textit{Note: this pass speculatively assumes that structured tensor operations are in place. If the assumption does not hold, the result will generally be sub-optimal since only a subset of code motion opportunities may be considered.}
\end{enumerate}
\end{Algo}

@@

In Figure~\ref{code:poisson}, Algorithm~\ref{algo:sharing-elimination} is executed in a realistic scenario, which originates from the bilinear form of a Poisson equation in two dimensions. We observe that $\mathbb{P} = \lbrace P^0, P^1\rbrace$, with $P^0 = \lbrace (z_0 a_{ik} + z_2 b_{ik}), (z_1 a_{ik} + z_3 b_{ik})\rbrace$ and $P^1 = \lbrace (z_0 a_{ij} + z_2 b_{ij}), (z_1 a_{ij} + z_3 b_{ij})\rbrace$. In addition, $|P^i| = |S_{P^i}| = 2$, so strategy (i) is applied to both partitions (step 3). We then have (step 4) $\mathbb{G} = (\lbrace t^0, t^1, t^2, t^3 \rbrace, \emptyset)$, although no rewrite rule is applicable since $\mathbb{S} = \mathbb{T}$.

@@

Pre-evaluation opportunities can be exposed through exploration of the expression tree transformation space. This would be challenging if we were to deal with arbitrary loop nests and expressions. We make use of a result -- the foundation of tensor contraction mode -- to simplify our task. As summarized in Section~\ref{sec:tc}, multilinear forms can be seen as sums of monomials, each monomial being an integral over the equation domain of products (of derivatives) of functions from discrete spaces. A monomial can always be reduced to the product between a ``reference'' and a ``geometry'' tensors. In our model, a reference tensor is simply represented by one or more sub-expressions independent of $L_e$, exposed after special transformation of the expression tree. This leads to the following algorithm.

\begin{Algo}[Pre-evaluation]
\label{algo:pre-evaluation}
\normalfont
Consider a finite element integration loop nest $\Lambda = [L_e, L_i, L_j, L_k]$. We dissect the normal form input expression into distinct sub-expressions, each of them representing a monomial. Each sub-expression is then factorized so as to split constants from $[L_i, L_j, L_k]$-dependent terms. This transformation is feasible\footnote{For reasons of space, we omit the detailed sequence of steps (e.g., expansion, factorization), which is however available at \url{https://github.com/coneoproject/COFFEE/blob/master/coffee/optimizer.py}}, as a consequence of the results in~\citeN{Kirby:TC}. These $[L_i, L_j, L_k]$-dependent terms are hoisted outside of $\Lambda$ and stored into temporaries. As part of this process, the reduction induced by $L_i$ is computed by means of symbolic execution. Finally, $L_i$ is removed from $\Lambda$. 
\end{Algo}

The pre-evaluation of a monomial introduces some critical issues:
\begin{enumerate}
\item Depending on the complexity of a monomial, a certain number $t$ of temporary variables is required if pre-evaluation is performed. Such temporary variables are actually $n$-dimensional arrays of size $S$, with $n$ and $S$ being, respectively, the arity and the extent (iteration space size) of the multilinear loop nest (e.g., $n=2$ and $S = J K$ in the case of bilinear forms). For certain values of ${<}t, n, S{>}$, pre-evaluation will shift the loop nest from a CPU-bound to a memory-bound regime, which may be counter-productive for actual execution time.
\item The transformations exposing $[L_i, L_j, L_k]$-dependent terms increase the arithmetic complexity of the expression (e.g., expansion tends to increase the operation count). This could outweigh the gain due to pre-evaluation.
\item A strategy for coordinating sharing elimination and pre-evaluation is needed. We observe that sharing elimination inhibits pre-evaluation, whereas pre-evaluation could expose further sharing elimination opportunities.
\end{enumerate}

@@

Since, for certain forms and discretizations, hoisting may bring the working set to exceed the size of some level of local memory (e.g. the last level of private cache on a conventional CPU, the shared memory on a GPU), we introduce the following \textit{memory constraints}.

@@

Note that Constraint~\ref{const:Le} reflects the policy decision that the compiler should not silently consume memory on global data objects. It has the effect of shrinking the transformation space. Constraint~\ref{const:TH} has both theoretical and practical implications, which will be carefully analyzed in the next sections.

\subsection{Completeness under factorization, code motion, and reduction pre-evaluation}
\label{sec:completeness}
We defined sharing elimination and pre-evaluation as high level transformations on top of basic operators such as code motion and factorization. Factorization addresses \textit{spatial redundancy}. The presence of spatial redundancy means that some operations are needlessly executed at two points in an expression. Code motion and reduction pre-evaluation, on the other hand, target \textit{temporal redundancy}; that is, the needless execution of the same operation with the same operands at two points in the loop nest. We remark that sharing elimination and pre-evaluation aim to minimize spatial and temporal redundancy relying on \textit{a specific set of operators}. Theoretically, lower operation counts may be found by exploiting other domain-specific properties, such as redundancies in basis functions. Constraint~\ref{const:Le} also limits the applicability of code motion, which further reduced the transformation space to be analyzed.

@@

The most trivial way of evaluating $\theta_{se}$ and $\theta_{pre}$ would consist of applying the actual transformations and simply count the number of operations. If, on one hand, this would be tolerable for $\theta_{se}$ (Algorithm~\ref{algo:sharing-elimination} tends to have negligible cost), the overhead would be unacceptable if we applied pre-evaluation -- in particular, symbolic execution -- to all bipartitions analyzed by Algorithm~\ref{algo:gamma}. We then seek an analytic way of determining $\theta_{pre}$.

The first step consists of estimating the \textit{increase factor}, $\iota$. This number captures the increase in arithmetic complexity due to the transformations exposing pre-evaluation opportunities. To contextualize, consider the example in Figure~\ref{code:increase_factor}. One can think of this as the (simplified) loop nest originating from the integration of a pre-multiplied mass matrix. The sub-expression \texttt{$f_0$*$b_{i0}$+$f_1$*$b_{i1}$+$f_2$*$b_{i2}$} represents the coefficient $f$ over (tabulated) basis functions (array $B$). In order to apply pre-evaluation, the expression needs be transformed to separate $f$ from all $[L_i, L_j, L_k]$-dependent quantities (see Algorithm~\ref{algo:pre-evaluation}). By product expansion, we observe an increase in the number of $[L_j, L_k]$-dependent terms of a factor $\iota = 3$.

@@

In general, however, determining $\iota$ is not so straightforward since redundant tabulations may result from common sub-expressions. Consider the previous example. One may add one coefficient in the same function space as $f$, repeat the expansion, and observe that multiple sub-expressions (e.g., $b_{10}*b_{01}*...$ and $b_{01}*b_{10}*...$) will reduce to identical tables. To evaluate $\iota$, we then use combinatorics. We calculate the $k$-combinations with repetitions of $n$ elements, where: (i) $k$ is the number of (derivatives of) coefficients appearing in a product; (ii) $n$ is the number of unique basis functions involved in the expansion. In the original example, we had $n=3$ (for $b_{i0}$, $b_{i1}$, and $b_{i2}$) and $k=1$, which confirms $\iota=3$. In the modified example, there are two coefficients, so $k=2$, which means $\iota=6$.

If $\iota \geq I$ (the extent of the reduction loop), we already know that pre-evaluation will not be profitable. Intuitively, this means that we are introducing more operations than we are saving from pre-evaluating $L_i$. If $\iota < I$, we still need to find the number of terms $\rho$ such that $\theta_{pre} = \rho \cdot \iota$. The mass matrix monomial in Figure~\ref{code:increase_factor} is characterized by the dot product of test and trial functions, so trivially $\rho = 1$. In the example in Figure~\ref{code:poisson}, instead, we have $\rho = 3$ after a suitable factorization of basis functions. In general, therefore, $\rho$ depends on both form and discretization. To determine this parameter, we look at the re-factorized expression (as established by Algorithm~\ref{algo:pre-evaluation}), and simply count the terms amenable to pre-evaluation.

\section{Formalization}
\label{sec:proof}
We demonstrate that the orchestration of sharing elimination and pre-evaluation performed by the transformation algorithm guarantees quasi-optimality (Definition~\ref{def:mln-quasi-optimality}). The proof re-uses concepts and explanations provided throughout the paper, as well as the terminology introduced in Section~\ref{sec:se-algo}.

\begin{Prop}
\label{prop:optimal-approach}
Consider a multilinear form comprising a set of monomials $M$, and let $\Lambda$ be the corresponding finite element integration loop nest. Let $\Gamma$ be the transformation algorithm. Let $X$ be the set of monomials that $\Gamma$ establishes to pre-evaluate, and let $Y = M \setminus X$. Assume that the pre-evaluation of different monomials does not result in identical tables. Then, $\Lambda' = \Gamma(\Lambda)$ is quasi-optimal in the sense of Definition~\ref{def:mln-quasi-optimality}, and satisfies Constraint~\ref{const:TH}.
\end{Prop}
\begin{proof}
We first observe that the cost function $\theta$ predicts the \textit{exact} gain/loss in monomial pre-evaluation, so $X$ and $Y$ can actually be constructed.

Let $c_\Lambda$ denote the operation count for $\Lambda$ and let $\Lambda_I \subset \Lambda$ be the subset of innermost loops (all $L_k$ loops in Figure~\ref{code:loopnest-opt}). We need to show that there is no other synthesis $\Lambda''_{I}$ satisfying Constraint~\ref{const:TH} such that $c_{\Lambda''_I} < c_{\Lambda'_I}$. This holds if and only if
\begin{enumerate}
\item \textit{The coordination of pre-evaluation with sharing elimination is optimal}. This boils down to prove that
\begin{enumerate}
\item \textit{pre-evaluating any $m \in Y$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
\item \textit{not pre-evaluating any $m \in X$ would result in $c_{\Lambda''_I} > c_{\Lambda'_I}$}
\end{enumerate}
\item \textit{Sharing elimination produces quasi-optimal (or optimal) loop nests.}\\
\end{enumerate}

@@

\begin{enumerate}
\item 
\begin{enumerate}
\item Let $T_m$ represent the set of tables resulting from applying pre-evaluation to a monomial $m$. Consider two monomials $m_1, m_2 \in Y$ and the respective sets of pre-evaluated tables, $T_{m_{1}}$ and $T_{m_{2}}$. If $T_{m_{1}} \cap\ T_{m_{2}} \neq \emptyset$, at least one table is assignable to the same temporary. $\Gamma$, therefore, may skip a potential global optimum, since $\theta$ only reasons on monomials in ``isolation''. We neglect this scenario (see assumptions) because of its purely pathological nature and its -- with high probability -- negligible impact on the operation count.
\item Let $m_1 \in X$ and $m_2 \in Y$ be two monomials that share some generic multilinear symbols. If $m_1$ were carelessly pre-evaluated, there may be a potential gain in sharing elimination that is lost, potentially leading to a local -- instead of a global -- optimum. This situation is prevented by construction, because $\Gamma$ analyzes all possible bipartitions searching for a global optimum that also preserves the memory constraints. Recall that since the number of monomials is in practice very small, this pass can rapidly be accomplished. Finally, note how this problem can be seen as an instance of the well-known Knapsack problem.
\end{enumerate}
\item Consider Algorithm~\ref{algo:sharing-elimination}. Because of normal form of expressions, loop linearity and structure of tensor operations, there are only two ways of scheduling the multilinear operands in $P \in \mathbb{P}$: through (i) generalized code motion or (ii) factorization of multilinear symbols. These two strategies lead, respectively, to performing $|P|$ or $|S_P|$ products at every loop iteration. Since (i) is applied if and only if $|P| < |S_P|$ and does not require neither expansion nor factorization, step (3) will not prune any optimal solution from the search space. %; in fact, it simply shifts towards the objective.

Steps (4)-(5) factorize symbols following the ordering dictated by the sharing graph and its rewrite rules. Recall that rules (A), (B), and (C) are applied in sequence, with applications of rule (C) that may expose further possibilities for (A) and (B). We prove that the application of these rules, in this sequence, represents a step towards quasi-optimality. 
\begin{description}
\item[Rule (A)] $\lbrace T^0, ..., T^n \rbrace$ can only be grouped through the factorization of $S$, so this rule is side effects free (i.e., no need to choose between different factorizations).
\item[Rule (B)] Since $S^{nt}$ is no terminal, there must exist at least one edge between $S^{nt}$ and a generic vertex $V$. Let us consider the alternative factorization $S^{nt}(...S + ... + ...V)$. If $V$ is a terminal, we are back to where we started, because rule (B) will be applicable by setting $V = T$ and $S = S^{nt}$. Otherwise ($V$ non terminal), we have $S$ appearing in two different sub-expressions, which implies sub-optimality because of the presence of spatial sharing.
\item[Rule (C)] We can interpret $G$ as a bipartite graph $\mathbb{B} = (\mathbb{S}_j, \mathbb{S}_k, \mathbb{E})$, in which $\mathbb{S}_j$ and $\mathbb{S}_k$ represent the partitions of symbols depending on $L_j$ and $L_k$, respectively. For each disconnected component $\mathbb{B^d} = (\mathbb{S}_j^d, \mathbb{S}_k^d, \mathbb{E}^d)$ in $\mathbb{G}$, we consider two possibilities, namely $\mathbb{B^d}$ 1) is not or 2) is complete. Case 1) is the simplest: since, at this point, there are no more hits for rules (A) and (B), the lower bound on the operation count (as number of products performed at every loop iteration), $lb$, is given by $lb = {min(}|\mathbb{S}_j^d|, |\mathbb{S}_k^d|{)}$. This is reachable by factorizing the vertex with highest degree (as the rule states), as any other choice may break factorization opportunities. In case 2), we have $lb = 1$ because factorization is applicable iteratively (e.g., $a_j a_k + a_j b_k + b_j a_k + b_j b_k = (a_j + b_j)(a_k + b_k) = t_j t_k$); this is handled by going back to step (2), at the end of step (5).
\end{description}
The algorithm is guaranteed to terminate because of the jump to step (6) when $\mathbb{S} = \emptyset$, before applying the rewrite rules. 
\end{enumerate}
\end{proof}
